### 2008-12-20 talk/ecoop2009/jitgen.tex
several small fixes.  Some rewording in the cli backend section

The JIT generation framework uses partial evaluation techniques to generate a
dynamic compiler from an interpreter; the idea is inspired by Psyco, which
uses the same techniques but it's manually written instead of being
automatically generated.

@@

The binding-time analyzer of our translation tool-chain is using a simple
abstract-interpretation based analysis. It is based on the
same type inference engine that is used on the source RPython program.
This is called the \emph{hint-annotator}; it
operates over input graphs that are already low-level instead of
RPython-level, and propagates annotations that do not track types but
value dependencies and manually-provided binding time hints.

@@

>>>>

The JIT generation framework uses partial evaluation techniques to generate a
dynamic compiler from an interpreter; the idea is inspired by Psyco \cite{DBLP:conf/pepm/Rigo04}, which
uses the same techniques but it's manually written instead of being
automatically generated.

@@

The binding-time analyzer of our translation tool-chain is using a simple
abstract-interpretation based analysis. It is based on the
same type inference engine that is used on the source RPython program.
This is called the \emph{hint-annotator}; it
operates over input graphs that are already low-level instead of
RPython-level, \anto{we never make distinction between low-level and rpython-level 
flowgraphs, do we? I propose to talk about ``intermediate flowgraphs''} 
and propagates annotations that do not track types but
value dependencies and manually-provided binding time hints.

@@

\anto{maybe we should at least mention promotion, and refer to the proper
  section for details?}

### 2008-12-20 talk/ecoop2009/rainbow.tex
several small fixes.  Some rewording in the cli backend section

Interpreters for dynamic languages typically continuously allocate a lot of small
objects, for example due to boxing. This makes arithmetic operations extremely
inefficient. For this reason, we
implemented a way for the compiler to try to avoid memory allocations in the
residual code as long as possible. The idea is to try to keep new
run-time instances "exploded": instead of a single run-time object allocated on
the heap, the object is "virtualized" as a set
of fresh local variables, one per field. Only when the object can be accessed by from
somewhere else is it actually allocated on the heap. The effect of this is similar to that of
escape analysis \cite{Blanchet99escapeanalysis}, \cite{Choi99escapeanalysis},
which also prevents allocations of objects that can be proven to not escape a
method or set of methods (the algorithms however are a lot more advanced than
our very simple analysis).

It is not always possible to keep instances virtual.  The main
situation in which it needs to be "forced" (i.e. actually allocated at
run-time) is when the pointer escapes to some non-virtual location like
a field of a real heap structure.  Virtual instances still avoid the run-time
 allocation of most short-lived objects, even in non-trivial situations.

@@

From a partial evaluation point of view, promotion is the converse of
the operation generally known as "lift" \cite{XXX}.  Lifting a value means
copying a variable whose binding time is compile-time into a variable
whose binding time is run-time – it corresponds to the compiler
"forgetting" a particular value that it knew about.  By contrast,
promotion is a way for the compiler to gain \emph{more} information about
the run-time execution of a program. Clearly, this requires
fine-grained feedback from run-time to compile-time, thus a
dynamic setting.

Promotion requires interleaving compile-time and run-time phases,
otherwise the compiler can only use information that is known ahead of
time. It is impossible in the "classical" approaches to partial
evaluation, in which the compiler always runs fully ahead of execution.
This is a problem in many realistic use cases.  For example, in an
interpreter for a dynamic language, there is mostly no information
that can be clearly and statically used by the compiler before any
code has run.

A very different point of view on promotion is as a generalization of techniques
that already exist in dynamic compilers as found in modern virtual machines for
object-oriented language.  In this context feedback
techniques are crucial for good results.  The main goal is to
optimize and reduce the overhead of dynamic dispatching and indirect
invocation.  This is achieved with variations on the technique of
polymorphic inline caches \cite{hoelzle_optimizing_1991}: the dynamic lookups are cached and
the corresponding generated machine code contains chains of
compare-and-jump instructions which are modified at run-time.  These
techniques also allow the gathering of information to direct inlining for even
better optimization results. 
\anto{What about this: While traditional PICs are only applied to indirect
  calls, promotion is a more general operation that can be applied to any kind
  of value, including instances of user-defined classes or integer numbers}

@@

Now, let us examine the content of the \lstinline{IntObj} case: first, there
is a hint to promote the class of \lstinline{b}.  Although in general
promotion is implemented through a flexswitch, in this case it is not needed
as \lstinline{b} holds a \emph{virtual instance}, whose class is already
known (as described in previous section).

Then, the compiler knows the exact class of \lstinline{b}, thus it can inline
the calls to \lstinline{lt}.  Moreover, inside \lstinline{lt} there is a
call to \lstinline{a.int_o()}, which is inlined as well for the very same
reason.

Moreover, as we saw in section \ref{sec:virtuals}, the \lstinline{IntObj}
instance can be virtualized, so that the subsequent \lstinline{BR_COND} opcode
can be compiled efficiently without needing any more flexswitch.

>>>>

Interpreters for dynamic languages typically continuously allocate a lot of small
objects, for example due to boxing. This makes arithmetic operations extremely
inefficient. For this reason, we
implemented a way for the compiler to try to avoid memory allocations in the
residual code as long as possible. The idea is to try to keep new
run-time instances \emph{exploded}: instead of a single run-time object allocated on
the heap, the object is \emph{virtualized} as a set
of fresh local variables, one per field. Only when the object can be accessed by from
somewhere else is it actually allocated on the heap. The effect of this is similar to that of
escape analysis \cite{Blanchet99escapeanalysis}, \cite{Choi99escapeanalysis},
which also prevents allocations of objects that can be proven to not escape a
method or set of methods (the algorithms however are a lot more advanced than
our very simple analysis).

It is not always possible to keep instances virtual.  The main
situation in which it needs to be \emph{forced} (i.e. actually allocated at
run-time) is when the pointer escapes to some non-virtual location like
a field of a real heap structure.  Virtual instances still avoid the run-time
 allocation of most short-lived objects, even in non-trivial situations.

@@

From a partial evaluation point of view, promotion is the converse of
the operation generally known as \emph{lift} \cite{XXX}.  Lifting a value means
copying a variable whose binding time is compile-time into a variable
whose binding time is run-time – it corresponds to the compiler
``forgetting'' a particular value that it knew about.  By contrast,
promotion is a way for the compiler to gain \emph{more} information about
the run-time execution of a program. Clearly, this requires
fine-grained feedback from run-time to compile-time, thus a
dynamic setting.

Promotion requires interleaving compile-time and run-time phases,
otherwise the compiler can only use information that is known ahead of
time. It is impossible in the ``classical'' approaches to partial
evaluation, in which the compiler always runs fully ahead of execution.
This is a problem in many realistic use cases.  For example, in an
interpreter for a dynamic language, there is mostly no information
that can be clearly and statically used by the compiler before any
code has run.

A very different point of view on promotion is as a generalization of
techniques that already exist in dynamic compilers as found in modern virtual
machines for object-oriented language, like \emph{Polymorphic Inline Cache}
(PIC, \cite{hoelzle_optimizing_1991}) and its variations, whose main goal is
to optimize and reduce the overhead of dynamic dispatching and indirect
invocation: the dynamic lookups are cached and the corresponding generated
machine code contains chains of compare-and-jump instructions which are
modified at run-time.  These techniques also allow the gathering of
information to direct inlining for even better optimization results. Compared
to PICs, promotion is more general because it can be applied not only to
indirect calls but to any kind of value, including instances of user-defined
classes or integer numbers.

@@

Now, let us examine the content of the \lstinline{IntObj} case: first, there
is a hint to promote the class of \lstinline{b}.  Although in general
promotion is implemented through a flexswitch, in this case it is not needed
as \lstinline{b} holds a \emph{virtual instance}, whose class is already
known (as described in the previous section).

Then, the compiler knows the exact class of \lstinline{b}, thus it can inline
the calls to \lstinline{lt}.  Moreover, inside \lstinline{lt} there is a
call to \lstinline{a.int_o()}, which is inlined as well for the very same
reason.  Moreover, as we saw in section \ref{sec:virtuals}, the \lstinline{IntObj}
instance can be virtualized, so that the subsequent \lstinline{BR_COND} opcode
can be compiled efficiently without needing any more flexswitch.

\anto{We should show the very final version of the code, with all instances virtualized}
